Research Experience

### 2010-03-02 doc/thesis/thesis.tex
little cleanup in bview. more thesis work.

\begin{document}
\section*{Introduction}
Functional Magnetic Resonance Imaging (FMRI) is a powerful tool in the analysis
neural activity. Despite its rather limited temporal resolution, FMRI still
is the best way of measuring neural activity deep in the brain. 
Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is relatively quick and cheap, and its analysis is straight forward.
Because of these benefits, FMRI continues to have great potential for
studying human cognition. Despite the fact that FMRI is so widely used,
the choice of analysis tools is still relatively limited. Every widely
available analysis tool is based on parametric modeling, which \emph{requires}
prior knowledge of the noise, and depends on that knowledge being accurate.
This limitation casts a long shadow over any experiment that claims to be
able to reject the null hypothesis, no matter the P-value. While obviously
there is a correlation between the BOLD signal and actual activity, its
possible the current parametric tests are over-estimating the actual amount
of activity. Its worht noting that the opposite could also be true, parametric
based t-tests may in fact underestimate activity.

@@

\subsection*{FMRI?}

\subsection*{BOLD Physiology}
The physiology that leads to the BOLD model has been well studied over the 
past decade and a half. While some uncertainty remains, for instance in
relation to the BOLD post-stimulus undershoot <papers relating to back and
forth on post-stimulus undershoot>, generally the model fits reality very 
well. As with every piece of tissue in the body, the brain requires oxygen
to extract energy from glucose in the blood. This process of removing oxygen
from red blood cells sets into motion a chain of events

inherently alters the local ratio of Oxygenated Hemoglobin
to De-oxygenated hemoglobin, which of course changes the atomic makeup of the
region. Of course MRI's contrast is based on the nuclear spin excitability of
atoms, and so change in the amount of oxygen changes the local oxygen concentration of 
The changes in the FMRI signal result from the changes in the ratio
of Oxygenated Hemaglobin to Deoxygenated Hemaglobin.

\section*{Current Techniques}
\subsection*{Basic T-Test}
\subsection*{General Linear Model}

>>>>

\begin{document}
\section*{Introduction}
Functional Magnetic Resonance Imaging (FMRI) is a powerful tool in the analysis
of neural activity. Despite its rather limited temporal resolution, FMRI is still
the best way of measuring neural activity deep in the brain. 
Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is relatively quick and cheap, and its analysis is straight forward.
Because of these benefits, FMRI continues to be crucial to the study of human 
cognition. Despite the fact that FMRI is so widely used,
the choice of analysis tools is still relatively limited. Every widely
available analysis tool is based on parametric modeling, which requires
prior knowledge of the noise distribution. Indeed, the linear methods
usually applied are known not to be robust to non-gaussian, non-white
noise, which is why so many pre-processing steps are necessary.
These limitations cast a long shadow over any experiment that claims to be
able to reject the null hypothesis, no matter the P-value. While obviously
there is a correlation between the BOLD signal and actual activity, its
possible the current parametric tests are over-estimating the actual amount
of activity. Its also possible that the current methods underestimate
actual activity. In this paper we propose a robust model-based approach to 
activation detection that makes no assumptions about the underlying noise
model.

@@

\subsection*{FMRI}
Before going further, it is necessary to discuss Functional MRI,
and what it measures. All MRI works by exciting nuclear spin away
from a steady state controlled by a large electromagnet. Different 
nuclei respond to RF impulses based on their natural frequency
of oscillation. Therefore contrast in MRI is based on the aggregate
nuclear composition of a given region. Thus, regions with more oxygen
will respond differently from regions with proportionately less 
oxygen. This exact difference is what causes the changes over time
in the MR signal in functional MRI. Unfortunately, it takes a significant
amount of time for nuclear spin to return to equilibrium. Thus instead
of waiting, a type of imaging called Echo Planar Imaging (EPI) is used,
wherein a single excitation is used, and then as quickly as possible
every voxel (3 Dimensional Pixel) in a particular region is read. 
Because "reading" a particular region requires the switching on of a
gradient magnetic field, it takes a decent amount of time to perform
measurements. Problematically, the signal from the original excitation
drops exponentially, thus the signal to noise ratio can be very low
even at modest resolutions. Eventually the signal becomes unreadable,
and a new excitation must be applied. As more measurements are made, and thus
as the spatial resolution increases, the rate at which entire volumes
can be acquired drops. Typically volumes can be acquired every two to
three seconds, at resolutions of about 64 by 64 by 32, but their signal
to noise ratio is often poor. Other imaging modalities such as EEG and
MEG are capable of much faster measurements, but are unable to isolate
measurement locations so precisely as FMRI. As with all biological systems,
noise is much higher than in human designed systems, and yet these 
biological systems are capable of performing amazing feats.

\subsection*{BOLD Physiology}
The physiology that leads to the BOLD model has been well studied over the 
past decade and a half. While some uncertainty remains (for instance in
relation to the BOLD post-stimulus undershoot <papers relating to back and
forth on post-stimulus undershoot>) generally the model fits reality very 
well. As with every piece of tissue in the body, the brain requires oxygen
to extract energy from glucose in the blood. This process of removing oxygen
from red blood cells sets into motion a chain of events that locally 
alters the composition of the blood. Inactive neurons can be thought of
as a slingshot cocked and ready to fire. As soon as a signal moves up
the axon and causes the neuron to fire (and thus changes the state
of a membrane at that location), ions quickly move across the altered
membrane to compensate for a high charge imbalance between the interior
and exterior of the cell. This process, similar to allowing a stretched
rubber band to contract leaves the system at a lower energy state. This cannot
last though, because the neuron needs to be ready to fire again rather quickly.
To return the cell to a ready position the membrane becomes impermeable
to the ions again, and then begins pumping ions back into the cell. This
process takes a large amount of energy, whereas the actual firing takes
very little energy. Thus, after firing, glucose is burned, removing oxygen
from the blood and thus causing a dip in the ratio of oxygenated hemoglobin to 
de-oxygenated hemoglobin. This is the first potentially measurable effect
that MRI can see, although it usually lasts fewer than 2 seconds making it
difficult for FMRI to catch. As a result of the decreased amount of local oxygen,
the capillaries compensate by increasing blood flow to that region. Because,
of the quickly increased blood flow into local capillaries, the blood
volume increases in addition to the local oxygen content. This leads to a Windkessel
effect which further lags the normalization of oxygen content. The effect of
all this is an overshoot in the oxygen content above the initial level. After
the work of recharging is done, there may be a prolonged undershoot lasting
as much as 90 seconds \cite{spatial_poststim}, though the reason for this
is debated \cite{origin_poststim}.

All these effects are generally accepted as the cause of the BOLD signal, 
but FMRI doesn't detect this happening in one neuron, but rather as the 
aggregate over hundreds or millions of cells. Though local neurons act
"together" (i.e. around the same time), the density of neurons, the
density of capillaries, and slight differences in activation across 
a particular voxel can all lead to signal attenuation or noise. 
A particularly insidious type of noise present in FMRI is a low frequncey
drift, characterized by a Weiner process. Though not present in all
regions, it is prevalent enough to cause problems \cite{detrend}. It is still not
clear where exactly this noise comes although it is possible it is 
the result of magnets heating up, or some distortion in magnetic
fields. It is clear that this drift signal is not due to a true physiological
effects however, given its presence in cadavers and phantoms\cite{drift}.

\section*{Current Techniques}
\subsection*{Basic T-Test}
The most basic method of analyzing FMRI data is through a basic T-test
between "resting state" and "active state" samples. This is done by 
taking the average and variance of the inactive period, and the 
active period separately then treating them both as gaussian distributions.
If they are in fact Gaussian distributions, then a basic t-test will
give the likelihood that the samples came from the same distribution
(the p-value). Of course, this test is fraught with problems; even if
the drift mentioned earlier has been removed, there is little reason
to believe that the noise is Gaussian, or even stable. Additionally, 
even if the noise were Gaussian, a t-test with a p-value of .05 over
5000 or more samples is on average going to generate $.05*5000$ false
positives. To compensate for this, bonferoni correction, also known as
multiple comparison tests are performed; essentially p-values are 
divided by the number of independent
tests being run. This, however, leads to extremely low p-values, so
low that it would be impossible for any biological system to satisfy. To
compensate, a Gaussian kernel is applied to the image, thus reducing
variance (and thus separating the active and inactive distributions)
as well as decreasing the effective number of voxels. Since t-tests are
now no longer being applied to n <I need to define n> independent voxels,
the factor by which the p-value must be divided by can be decreased.
<Do I need to mathematically define all this?> The derivation and application
of random field theory, and its use can be found in various papers \cite{univ_mult_rft}.

\subsection*{General Linear Model}

\subsection*{Whats wrong with these techniques}
There are a few problems with the techniques mentioned in the 
previous sections. First, they essentially ignore prior knowledge about
the system. The physiological models for the BOLD signal are quite good
and based in physics. Thus the parameters may be constrained by reality,
which crucially reduces model variance. Additionally, lacking real world
parameters means that the parameters have no clear physiological meaning,
so they are less useful in identifying pathologies.

Secondly, these methods are still merely t-tests, which notoriously
lack robustness to non-Gaussian noise. While different techniques 
exist for imposing Gaussianity, those techniques are incapable of
discriminating noise from signal, otherwise analyzing FMRI would be
relatively simple. There is no way to know how much signal is removed
by various smoothing techniques, or even if entire regions have been
smoothed into oblivion. Instead of extensively filtering data to 
remove noise, the analysis method itself must be robust to noise,
which is why we propose here the use of particle filters.
